---
title: "Week 7 lab"
author: "Saransh Rakshak"
date: "10/22/2024"
output:
  html_document: default
---
  
```{r setup, include=FALSE}
library(dplyr)
```

(You might want to knit the first time you read this document to view the notation.)

# Inference on one proportion using R

## Example (defective widgets)

An engineer wants to learn the percentage of defective widgets that come out of the production line. The plant manager believe that 3% are defective, but she believes it might be higher. A simple random sample of 320 widgets is selected one day and the number of defects is counted. 

Instead of giving you a fictional data set to download, I am going to just simulate data for this example. The code below produces some binary data that could arise in this scenario. (In this simulation, the engineer is correct and the true proportion is 0.04, not 0.03.)  The variable `defective` will equal 0 if the widget is not defective and 1 if it is.  

```{r}
# generate a data set of simulated widgets
# 
n <- 320  # number of widgets to sample from the line
true.pi <- 0.04  # true proportion of defective widgets
defects <- rbinom(n, 1, prob = true.pi) # indicator of whether it is defective

# put it all into a data frame
widget_data <- data.frame(widget_no = 1:n, defective = defects)
head(widget_data)
```

## Descriptive analysis
First, I'll summarize the `defect` variable using the `table` and `prop.table` functions.
```{r}
table(widget_data$defective)
prop.table(table(widget_data$defective))
```

```{r}
head(widget_data)
```

## Confidence interval for the proportion
To make a 90% confidence interval for pi, the true proportion of widgets that are defective, I can use the `prop.test` function as shown below. 

```{r}
successes <- sum(widget_data$defective == 1)
n <- nrow(widget_data)
prop.test(successes, n, conf.level=0.90)
```


By default, this function prints a lot of information.

If you are only interested in the confidence interval, you can extract it by appending `$conf.int` to the call of `prop.test`.  

```{r}
ci_pi <- prop.test(successes, n, conf.level=0.90)$conf.int
ci_pi
```

Remember that the default settings in R use a continuity correction and a small-sample correction that will not give the exact same results as the large-sample intervals discussed in lecture.  


## Testing for pi
The same `prop.test` function performs a hypothesis test for $\pi$.  By default, it tests the hypotheses $$H_0:\pi=0.5; \quad H_A: \pi \neq 0.5.$$

You can change the null value ($\pi_0$) using the `p` option. You can change the alternative using the `alternative` option (the options are "two.sided", "less", and "greater").  The option `correct=FALSE` will result in the large-sample test from lecture being performed.

We will now test our hypothesis, $$H_0:\pi=0.03; \quad H_A: \pi > 0.03.$$
```{r}
test_result <- prop.test(successes, n, p=0.03, alternative='greater', correct=FALSE)
print(test_result)
```

You can extract the p-value for the test as follows:
```{r}
test_result$p.value
```

You can also find the absolute value of the $z_0$ statistic using the code below. It's good practice to report this value as well as the p-value.

```{r}
z.0 <- sqrt(test_result$statistic)
print(z.0)
```

# Exercises


## Ex 1: Poisonous mushrooms

The file `mushrooms.csv`, posted on Canvas in the Data module, contains records that represent a simple random sample of mushroom species from the Agaricus and Lepiota family. Today we'll look at the variable `edible`, which is equal to "e" if the species is edible and "p" if it is toxic.  


a. How many species in the sample are edible and how many are poisonous?  What proportion are edible?

**Answer: Edible = 219, Poisonous = 206, Proportion Edible = 0.5152941**

```{r ex1a}
mushrooms <- read.csv('mushrooms.csv')
table(mushrooms$edible)
prop.table(table(mushrooms$edible))
```


b. Calculate a 99% large-sample confidence interval for the proportion of mushrooms that are edible. Use the formula for a large-sample interval given in lecture.

**Answer: Conf Interval = (45.28%, 57.77%)** 

```{r ex1b}
pi.hat <- sum(mushrooms$edible == 'e') / nrow(mushrooms)
z.star <- qnorm(1 - 0.01 / 2)
std.error <- sqrt(pi.hat * (1 - pi.hat) / nrow(mushrooms))

lower_bound <- pi.hat - z.star * std.error
upper_bound <- pi.hat + z.star * std.error

print(paste(lower_bound, upper_bound))
```


c. Use the `prop.test` function to find a 99% confidence interval for the proportion of mushrooms that are edible.  Does it differ from the one you found in part b?  If so, is the difference substantial enough to lead you to different conclusions? 

**Answer: prop.test Conf Interval = (45.31%, 57.70%) This change in confidence interval is not enough to lead to a new conclusion.**

```{r ex1c}
edible_cnt <- nrow(mushrooms %>% filter(mushrooms$edible == 'e'))
n_mush <- nrow(mushrooms)
conf_lvl <- 0.99

prop.test(edible_cnt, n_mush, conf.level = conf_lvl, correct = FALSE)
```

d. Use the `prop.test` function in R to test whether the data provide strong evidence that the proportion of edible species is not equal to 0.5.  Report the test statistic, p-value, decision, and summarize your conclusion. Use $\alpha=0.01$.

**Answer: The calculated test statistic is 0.5820855 and p-value is 0.5605. Since our p-value is greater than $\alpha=0.01$ we fail to reject the null hypothesis (i.e. fail to reject $\pi=0.5$). We can conclude that the data does not provide sufficient evidence to conclude that the proportion of edible mushroom species is not 0.5.**

Our hypotheses: $$H_0:\pi=0.5; \quad H_A:\pi\neq 0.5$$

```{r ex1d}
edible_cnt <- sum(mushrooms$edible == 'e')
n_mush <- nrow(mushrooms)
pi <- 0.5

test1 <- prop.test(edible_cnt, n_mush, p = pi, alternative = 'two.sided')
test1

z.0 <- sqrt(test1$statistic)
z.0
```


## Ex 2: Mushroom gills

Estimate (make a confidence interval for) the proportion of mushrooms whose `gill.size` is broad. (For this variable, b=broad; n=narrow.) Report and interpret the interval. Use a reasonable confidence level of your choosing.

**Answer: I can conclude that I am 95% confident between 63.1% and 72.1% mushrooms have gill size of "broad".**

```{r ex2}
broad_gilled_cnt <- sum(mushrooms$gill.size == 'b')
n_mush <- nrow(mushrooms)
conf_lvl <- 0.95

gill_test <- prop.test(broad_gilled_cnt, n_mush, conf.level = conf_lvl)
gill_test

z.0 <- sqrt(gill_test$statistic)
z.0
```

## Ex 3: colored mushrooms
a. Look at the mushrooms for whose `cap.color` equal `w` (white). Test whether there is evidence that the proportion of these mushrooms that are edible is greater than 0.5. Use $\alpha=0.05$. 

Report the test statistic, p-value, and conclusion.

Our Hypothesis: $$H_0:\pi=0.5; \quad H_A: \pi > 0.5.$$

**Answer: We find the test statistic to be $|z_0|=3.341$ and the p-value = 0.0004177. Since our p-value is greater than $\alpha=0.05$ we can reject the null hypothesis (i.e. reject $\pi=0.5$). Thus we can conclude that we are 95% confident that more than 50% of white mushrooms are edible.**

```{r ex3asumry}
white <- mushrooms %>% filter(cap.color == 'w')

table(white$edible)
prop.table(table(white$edible))
```

```{r ex3ahypotest}
white_edible_cnt <- sum(white$edible == 'e')
white_cnt <- nrow(white)
pi <- 0.5

white_test <- prop.test(white_edible_cnt, white_cnt, p = pi, alternative = 'greater')
white_test

z.0 <- sqrt(white_test$statistic)
z.0
```


b. Compare your findings to exercise 1. Do your findings suggest any association between cap color and edibility? 

**Answer. In the previous exercise, we considered the all mushrooms regardless of color and concluded that the data gave no strong evidence that more than 50% of mushrooms are edible. In this exercise we found strong evidence that more than 50% of white mushrooms are edible. Thus we can see some association between mushroom color and edibility. **



## Ex 4: Writing a function (extra practice -- not required)

a. Write an R function to calculate the large-sample confidence interval for $\pi$.  Revisit Lab 1 for an example of writing a function.

**Answer.**

```{r ex4}
# Function for confidence interval using lecture formula
conf_int <- function(n, x, alpha) {
  p <- x / n
  z_score <- qnorm(1 - alpha / 2, 0, 1)
  z <- sqrt((p * (1 - p)) / n) * z_score
  interval <- c(p - z, p + z)
  return(interval)
}

conf_int(200, 100, 0.01)
```

```{r ex4b}
# Function for confidence interval using lecture formula
conf_int2 <- function(n, x, alpha) {
  prop <- x / n
  z.star <- qnorm(1 - alpha / 2)
  std.error <- sqrt(prop * (1 - prop) / n) * z.star
  lower_bound <- prop - std.error
  upper_bound <- prop + std.error
  return(c(lower_bound, upper_bound))
}

conf_int2(200, 100, 0.01)
```

```{r ex4c}
# Function for confidence interval using prop.test()
conf_int3 <- function(n, x, alpha) {
  test_result = prop.test(x, n, conf.level = 1 - alpha, correct = FALSE)
  return(test_result$conf.int)
}

conf_int3(200, 100, 0.01)
```







