---
title: "Week 5 lab"
author: "your name here"
date: "Completion date"
output:
  html_document: default
  pdf_document: default
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Normal quantile plots

Very often in statistics we assume that observed data values are realizations of random variables from a normal distribution.  A normal quantile plot is used to check whether that assumption is reasonable.  

If data are from a perfect normal distribution, the normal quantile plot will show points along a perfect 45-degree, uphill line.  The code below will generate random normal variables and display the quantile plot.

```{r}
x1 <- rnorm(5e3) # simulates random normal variables
qqnorm(x1)
qqline(x1, col='red')  # adds the line that perfectly normal data would follow
```

With a smaller sample, the plot will not look as perfect, but the general linear pattern will be the same. 

```{r}
x2 <- rnorm(15)
qqnorm(x2)
qqline(x2, col='red')  
```

If data are not normal, you may see an S-shaped pattern in the quantile plot, or curvature in the lower or upper tail.  The chunks below generate data from some skewed distributions and show their quantile plots.


```{r}
par(mfrow=c(1,2))
x3 <- rgamma(5e3,1,10)  # samples from a very skewed gamma distribution
hist(x3, main='x3 - not normal') 
qqnorm(x3)
qqline(x3, col='red')
```

```{r}
par(mfrow=c(1,2))
x4 <- rgamma(5e3, 10,10) # samples from a nearly symmetric gamma distribution
hist(x4, main = 'x4 - not normal') 
qqnorm(x4)
qqline(x4,col='red')
```

```{r}
par(mfrow=c(1,2))
x5 <- rpois(5e3,4) # samples from a discrete distribution
hist(x5, main='x5 - not normal \n discrete') 
qqnorm(x5)
qqline(x5,col='red')
```

In some of these examples, the histograms clearly indicate that the shape of the distribution does not resemble a normal curve.  Sometimes, however, the quantile plot reveals patterns that are harder to spot in a histogram.  Here is a sample from a t distribution, which is bell-shaped but has fatter tails than a Normal curve. The non-normality is much easier to spot in the normal quantile plot.

```{r}
par(mfrow=c(1,2))
x6 <- rt(5e3,6)
hist(x6, breaks=20)
qqnorm(x6)
qqline(x6,col='red')
```

## Exercises

## Exercise 1: interpreting probability density functions
Run the code chunk below to see a plot of a "Exponential" probability density function. Then answer the questions using just the plot. (You are not expected to know the exact answers; however, you should be able to form reasonable guesses.)


```{r, plot-exponential}
x.seq <- seq(0,15,by=0.01) # creates a grid of x-values at which to evaluate p(x).
plot(x.seq, dexp(x.seq,rate=0.5),type='l',main='Exponential distribution with rate=1/2', xlab='x', ylab='p(x)')
```

a. Based on the plot, give an approximate guess of P(X < 5).

b. Based on the plot, which probability is bigger: P(1 < X < 4) or P(3 < X < 6)?

c. Based on the plot, what do you think the 95th percentile of the distribution is?

d. Based on the plot, what do you think the 50th percentile of the distribution is?

e. Which is bigger: the 45th percentile of the distribution, or the 65th percentile?


## Exercise 2: Normal probabilities
Use the functions `pnorm()` and `qnorm()` to find the following probabilities.

a. Find the 77th percentile of a Normal distribution with mean 0 and standard deviation of 0.5.

**Answer: **

b. Find the probability that a N(0, 0.5^2) random variable is less than -0.35.

**Answer: **

c. Find the probability that a N(10, 3^2) random variable is greater than 17.

**Answer: **

d. Find the probability that a N(10, 3^2) random variable is between 9 and 14.

**Answer: **


## Exercise 3: Normal probabilities
The age of employees at a certain company have a Normal(37,4.7^2) distribution.

a.   Find the probability that a randomly selected employee is older than 40.

**Answer: **

b.  Find the cutoff for the youngest 10% of employees; that is, 10% of employees are younger than what age?

**Answer: **

c. Find the probability that a randomly selected employee is between 33 and 39.

**Answer: **

d.  Find the cutoff for the oldest 5% of employees; that is, 5% of employees are older than what age?

**Answer: **

## Exercise 4: simulation to approximate probabilities (optional)
Simulation can be used to find approximate probabilities.  It can also be used to find approximate distributions.  If you take a large number of samples (no less than 500) from a probability distribution, the histogram of the samples will look very similar to the true probability density function (or pmf). 

a.  The `rnorm()` function can be used to take samples from a normal distribution.  Here is an example of using `rnorm` to draw n=5 samples from a N(-4,0.8^2) distribution.

```{r}
y <- rnorm(n=5, mean=-4, sd=0.8)
print(y)
```

Now use `rnorm` to draw 1000 samples from the N(21.4, 2.8^2) distribution.  (Do not print the samples.)  Make a histogram of the samples. Does the shape of the histogram resemble a normal curve? Then calculate the mean and standard deviation of the samples -- do these numbers come close to the true mean and standard deviation?

**Answer: **

b.  Using the samples from part (a), calculate the proportion of samples that are greater than 25.  

**Answer: **



c. Use `pnorm()` to find the probability that a N(21.4, 2.8^2) random variable is greater than 25. Your answer in part (b) is an approximation of this probability. 

**Answer: **

## Exercise 5: gamma simulation to approximate probabilities (optional)

The code below gives an example of drawing random samples from a Gamma distribution using the `rgamma` function. The code draws n=5 samples from the Gamma distribution with shape parameter equal to 0.5 and scale parameter equal to 1.  (This is sometimes called the Gamma(0.5,1) distribution.) 

```{r}
y <- rgamma(5, shape=0.5, scale=1)
print(y)
```

a. Modify the code to generate n=10000 samples. (Do not print them.) Create a histogram and comment on the shape of the distribution.  This shape will be similar to the true density curve of the Gamma(0.5,1) distribution.

**Answer: **

b. Using the samples you obtained in part a, approximate the probability that a Gamma(0.5, 1) random variable is between 1 and 2.

**Answer: **

## Exercise 6: Quantile plots

a. Use the `rnorm` function to generate a sample of size 100 from a Normal(4, 0.75^2) distribution.  Create a normal quantile plot of the sample and state what features of the plot indicate whether it is reasonable to consider the sample to be approximately normal.

**Answer: **

b. Use the `rgamma()` function to generate a sample of size 75 from a Gamma distribution with shape=0.5 and scale=1.  Create a normal quantile plot of the sample and state what features of the plot indicate whether it is reasonable to consider the sample to be approximately normal.

**Answer: **

