---
title: "Week 4 lab"
author: "Suggested solution"
date: "submission date"
output:
  html_document: default
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Exercises


## Exercise 1: Poisson calculations 

Suppose the random variable $Y$ has a Poisson distribution.  Compute the following probabilities

a. $P(Y=4)$ given $\lambda=2$.

**Answer:**
```{r}
lambda.value <- 2
dpois(4,lambda.value)
```

b. $P(Y=4)$ given $\lambda=3.5$.

**Answer:**
```{r}
lambda.value <- 3.5
dpois(4,lambda.value)
```

c. $P(Y>4)$ given $\lambda=2$.

**Answer:**
```{r}
lambda.value <-2
ppois(4,lambda.value, lower.tail=FALSE)
```

d. $P(1 \leq Y < 4)$ given $\lambda=2$.

**Answer:**
```{r}
lambda.value <-2
dpois(1,lambda.value) + dpois(2,lambda.value) + dpois(3,lambda.value)
```

or

```{r}
lambda.value <-2
ppois(3,lambda.value) - ppois(0,lambda.value)
```

## Exercise 2: quality control 

The quality control department examines all the products returned to a store by customers.  An examination of the returned products yields the following assessment: 5% are defective and not repairable, 45% are defective but repairable, 35% have small surface scratches but are functioning properly, and 15% have no problems. 

Compute the following probabilities for a random sample of 20 returned products.  (Hint: frame the events as Binomial experiments, with differing values of $\pi$.)

**Answer:**
I will assume from now on that a "problem" includes scratches and defects and that any "defect" is not functioning properly. 


a. All of the 20 returned products have some type of problem.
```{r}
n <- 20
pi.value <- 0.05 + 0.45 + 0.35
dbinom(20,size=n, prob=pi.value)
```

b. Exactly 6 of the 20 returned products are defective and not repairable.

```{r}
pi.value <- 0.05 
dbinom(6,size=n, prob=pi.value)
```

c. Of the 20 returned products, 6 or more are defective and not functioning properly.

```{r}
pi.value <- 0.05 +0.45
# note: P( X >= 6) = P(X > 5) since X is discrete
pbinom(5,size=n, prob=pi.value, lower.tail=FALSE)
```

d. None of the 20 returned products has any sort of defect.

```{r}
pi.value <- 0.05 + 0.45
dbinom(0,size=n,prob=pi.value)
```


## Exercise 3: expected value of a binomial random variable
For parts a, b, and c, assume that the random variable X has a binomial distribution with n=8 and $\pi$=0.3.

a. Using information from the Binomial portion of lecture this week, what is the expected value (mean) of X?

**Answer:**
The mean is n*pi, which here is equal to 2.4.
  
b.  The answer will be the same if you use the general formula for expected value, 

$$E(X) = \sum_{x\in\mathcal{X}}xP(X=x). $$
  
Write code below that will calculate E(X) according to this formula. Find the $P(X=x)$ values using either the formula for the binomial pmf from lecture or the `dbinom()` function.  Your answer should match part a.

**Answer:**
Here is one way to calculate the expression.
```{r}
E.X <- 0*dbinom(0,8, .3) + 1*dbinom(1,8, .3) + 2*dbinom(2,8, .3) + 3*dbinom(3,8, .3) +
  4*dbinom(4,8, .3) + 5*dbinom(5,8, .3)+ 6*dbinom(6,8, .3) + 7*dbinom(7,8, .3) +
  8*dbinom(8,8, .3)
E.X
```

If you like shortcuts, here is one.  This is an example of using "vectorized" operations in R.

```{r}
X.values <- seq(0,8, by=1)
X.values

# calculate a vector of probabilities for every value in X.values
dbinom(X.values, size=8, prob=0.3)

# Find expected value
sum(X.values*dbinom(X.values, size=8, prob=0.3))
```

c.  Use the `rbinom()` function to approximate the E(X) through simulation.  Try taking samples of size of 100, 1000, 5000, and 10,000.  How many samples does it take to obtain what appears to be a reasonably close approximation?

**Answer:**
Here is my simulation for a sample size of 100.
```{r}
n.samples <- 100
samples <- rbinom(n.samples,size=8,prob=0.3)
mean(samples)
```

The approximation is not equal to 2.4, but it is somewhat close. I can change "n.samples" to try the approximation for larger samples and find that the approximation tends to improve as the number of samples increases.

To summarize the differences, I'm going to save the results for different sample sizes. The following code uses sample sizes of 1000, 5000, and 10,000. For each, I store the approximations of E(X) in a vector and then print and plot the results.

```{r}
n.values <- c(100,1000,5000,10000)  # vector of n values
E.X.estimates <- rep(NA, length(n.values)) # vector to store the E(X) estimates

for(j in 1:length(n.values))
{
  temp.samples <- rbinom(n.values[j], size=8,prob=0.3)
  E.X.estimates[j] <- mean(temp.samples)
}

E.X.estimates
```
Here is a plot of the approximations. The horizontal line represents the true value of 2.4.
```{r}
plot(n.values, E.X.estimates, main="Approximations of E(X)",xlab="n samples")
abline(h=2.4,col="navy",lwd=1.4)
```

Based on the estimates, it seems as though 1000 samples is typically enough to obtain a good approximation to the true value of E(X)=2.4.  When I use 100 samples, the approximation is often pretty far from the true value, so I would say that 100 is not sufficient.


## Exercise 4: expected value of a Poisson random variable 

Use the `rpois()` function to approximate E(Y) if Y is a Poisson random variable with $\lambda=8.2$.

**Answer:**
The code below performs the calculation.
```{r}
poisson.samples <- rpois(10000,lambda=8.2)
E.y <- mean(poisson.samples)
E.y
```


## Exercise 5: unknown pi
I generated three random variables from a Binomial distribution with $n=30$ and a fixed value of $\pi$.  The values are 6, 8, and 5.  

The value of $\pi$ is unknown to you, but I will tell you it is one of the following values: 0.17, 0.24, 0.31, 0.39.  Use the `dbinom()` function to find $P(X=6)$, $P(X=8)$, and $P(X=5)$ for each of the possible values of $\pi$:

a. $\pi=$ = 0.17

b. $\pi$ = 0.24

c. $\pi$ = 0.31

d. $\pi$ = 0.39

Do the calculations give you any insight into which $\pi$ values are most plausible?  Give a guess as to which $\pi$ value is the one I used to generate the random sample.

**Answer:**
Answers will vary here. My approach is to calculate the probability of X=6, 8, and 5 for each of the pi values. The code chunks below do that.

a. $\pi=$ = 0.17

```{r}
X.values <- c(6,8,5) # values of X
dbinom(X.values, size=30,prob=0.17) # calculate probabilities for all 3 values
```

b. $\pi$ = 0.24

```{r}
dbinom(X.values, size=30,prob=0.24)
```

c. $\pi$ = 0.31

```{r}
dbinom(X.values, size=30,prob=0.31)
```

d. $\pi$ = 0.39

```{r}
dbinom(X.values, size=30,prob=0.39)
```

Based on these calculations, I do not think that 0.39 is a very plausible value for pi because each of the sampled values have probability of less than 0.06 when pi is 0.39.  Of the potential values of pi, the value X=8 is most probable when pi=0.24. Both X=6 and X=5 are most probable under pi=0.17.  Because of this, I would say that 0.17 and 0.24 are the most plausible values of pi. 

