---
title: "Week 9 Lab"
author: "your name here"
date: "Submission date"
output:
  html_document: default
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

(You might want to knit the document in orders to view the notation.)


# Inference on two means 

Before beginning an analysis that compares two means, you might want to think through these questions:

1. Are the data paired or independent samples?
2. Would a test or a confidence interval be a more appropriate approach?
3. If the data are two independent samples, is it more reasonable to assume an equal variances or unequal variances model?

When it comes to question 2, in many practical applications it might be your choice.  Testing is great if your primary interest is in whether the data provide strong evidence that the means of the two groups is not equal to a single, specific value.  If you simply want to learn about how different the means are, a confidence interval will provide more information.

Also, remember that a hypothesis test can be viewed as an "inversion" of a confidence interval.  In the context of inference on two means, this means that if the $(1-\alpha)\cdot 100$% CI for $\mu_1 - \mu_2$ includes some value, $D_0$, then we would fail to reject $H_0: \mu_1-\mu_2 = D_0$ against the two-sided alternative. The CI and the test ultimately provide the same information. 

In R, if you want to use a $D_0$ value other than 0 for a two independent sample t test, use the "mu=" option.  Here's an example:
```{r}
# simulate y1 and y2 from two normal distributions
n1 <- 7; n2 <- 9 # sample sizes in each group
y1 <- rnorm(n1,2.5,1)
y2 <- rnorm(n2,1,1)

# test H0: mu1 - mu2 = 1 vs HA: mu1-mu2 =/= 1
t.test(y1,y2,mu=1, var.equal=FALSE)

```



# Ex 1 (diabetes) 
National Institute of Diabetes and Digestive and Kidney Diseases conducted a study on diabetes. A population of several hundred Pima Indian women living near Phoenix, Arizona were tested for diabetes. Other information was gathered from these women at the time of testing, including number of pregnancies, glucose level, blood pressure, skin fold thickness, body mass index, diabetes pedigree and age. A random sample (n=44) of the data from this study are contained in the file `diabetes_sm.csv`.

a. Compare the ages of women who tested negative for diabetes and those who tested positive by making side-by-side boxplots of ages in the two groups.  Also compare the mean, median, and standard deviation of "age" across the two groups.  Use the variables "age" and "diabetes." Based on these descriptive summaries, does age appear to be associated with result of the diabetes test?

**Answer.**


b.  Let $\mu_1$ represent the mean age among women who tested positive for diabetes and let $\mu_2$ represent the mean age among women who tested negative.  Find a 95% confidence interval for $\mu_1 - \mu_2$.  

**Answer.**


c.  Using the same definitions of $\mu_1$ and $\mu_2$ as in part (b), test the hypotheses
$H_0: \mu_1-\mu_2 = 0; H_A: \mu_1 - \mu_2 \neq 0.$  Use $\alpha=0.05$.

(Before you perform the test, use the confidence interval in part (b) to predict in your mind whether $H_0$ will be rejected.)
Report the test statistic, p-value, and conclusion.

**Answer.**



# Ex 2: (tomato plants)

The file `tomato.csv` contains data on the heights of ten
tomato plants (in cm), grown under a variety of soil pH conditions and randomly sampled from their respective plots. Each plant was measured twice. During the first measurement, each plant’s height was recorded and a reading of soil pH was taken. During the second measurement only plant height was measured, because it is assumed that pH levels did not vary much from measurement to measurement.

A researcher wants to learn how much the plants grew on average between the first and second measurements.  

a. Provide a short descriptive analysis that addresses the researcher's question. (Make a relevant plot and calculate a few relevant summary statistics.)

**Answer.**


b. Find a 90% confidence interval for the average growth between the first and second measurements.

**Answer.**

c. Check the assumptions of the statistical model using a normal quantile plot.  Write a sentence stating whether the model seems to be a good approximation for the data.

**Answer.**


# Ex 3 (oxygen)

The Department of Natural Resources received a complaint from recreational fisherman that a community was releasing sewage into the river where they fish.  These types of releases lower the level of dissolved oxygen in the river and hence cause damage to the fish residing in the river.  An inspector from the DNR collective fifteen water samples from locations on the river upstream from the community and fifteen samples are selected from locations downstream from the community.  The dissolved oxygen readings in parts per million (ppm) were recorded at each location. 

The R code below reads the measurements into two objects, "upstream" and "downstream".  Use these for the exercise.  

```{r}
upstream <- c(5.2, 4.8, 5.1, 5,4.9, 4.8,5,4.7, 4.7, 5,4.6,5.2,5,4.9,4.7)
downstream <- c(3.2,3.4,3.7,3.9,3.6,3.8,3.9,3.6,4.1,3.3,4.5,3.7,3.9,3.8,4.5)
```
a. Are the upstream and downstream measurements paired or independent samples? 

**Answer.**


b. Make a plot to visually investigate whether the dissolved oxygen readings tend to be lower in the downstream or upstream locations.

**Answer.**


c. The DNR will intervene if the average dissolved oxygen levels upstream is more than 0.5ppm greater than the average dissolved oxygen levels downstream.  Do the data provide strong evidence that the difference in means is greater than 0.5?  Report the hypotheses, test statistic, p-value, and conclusion.

**Answer.**


d.  Check the normality assumptions of the statistical model using quantile plots. Write a sentence stating whether the model seems to be a good approximation for the data.


**Answer.**



# Ex 4 (more tomatos)

Revisit the tomato data in Ex 2.  

a. Find a 95\% confidence interval for the mean pH level for the population of tomatoes.

**Answer.**



# Ex 5 (more diabetes)

Revisit the diabetes data.  


a. Provide a descriptive plot and summary statistic that describes the relationship between glucose (glu) and diastolic blood pressure (bp).  Does there appear to be a strong association between the two variables?

**Answer.**


b. Create a binary variable that indicates whether an observation has `npreg` (number of pregnancies) greater than 3. Is there an association between this new variable and `diabetes`? Use descriptive methods and/or inferential methods from last week to answer the question.

**Answer.**


