---
title: "Week 7 Lab suggested solution"
author: "your name here"
date: "Submission date"
output:
  html_document: default
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




# Exercises


## Ex 1: Poisonous mushrooms

The file `mushrooms.csv`, posted on Canvas in the Data module, contains records that represent a simple random sample of mushroom species from the Agaricus and Lepiota family. Today we'll look at the variable `edible`, which is equal to "e" if the species is edible and "p" if it is toxic.  


a. How many species in the sample are edible and how many are poisonous?  What proportion are edible?

**Answer:**
The sample contains 219 edible mushrooms and 206 poisonous. The proportion of edible mushrooms is 0.51529.

```{r}
mush <- read.csv("mushrooms.csv",header=TRUE)
table(mush$edible)
prop.table(table(mush$edible))
```

b. Calculate a 99% large-sample confidence interval for the proportion of mushrooms that are edible. Use the formula for a large-sample interval given in lecture.

**Answer:** 
The large sample confidence interval is (0.4528503, 0.5777379). See the code below for the calculation. We are 99% confident that between 45.29% and 57.78% of mushrooms are edible.

```{r}
n <- nrow(mush)
pi.hat <- sum(mush$edible=='e')/n
z.star <- qnorm(1-0.01/2)
std.error <- sqrt(pi.hat*(1-pi.hat)/n)

pi.hat - z.star*std.error
pi.hat + z.star*std.error

```

c. Use the `prop.test` function to find a 99% confidence interval for the proportion of mushrooms that are edible.  Does it differ from the one you found in part b?  If so, is the difference substantial enough to lead you to different conclusions? 

**Answer:**
The code below finds the 99% CI to be (0.4519339, 0.5781753). This is different than the large-sample itnerval from part b, due to the continuity and sample size corrections discussed in lecture. The differences are not substantial enough to be of concern on a practical level.

```{r}
prop.test(sum(mush$edible=='e'),n,conf.level=0.99)
```


d. Use the `prop.test` function in R to test whether the data provide strong evidence that the proportion of edible species is not equal to 0.5.  Report the test statistic, p-value, decision, and summarize your conclusion. Use $\alpha=0.01$.

**Answer:**
The hypotheses are $H_0:\pi=0.5$; $H_A:\pi\neq 0.5$. The test statistic, calculated in the code below, is $|z_0|=0.582$ and the p-value is 0.5605. The p-value is greater than alpha, so fail to reject the null and conclude that the data provide insufficient evidence to conclude that the proportion of edible species is not 0.5.

```{r}
test1 <- prop.test(sum(mush$edible=='e'),n,p=0.5, alternative='two.sided')
test1

sqrt(test1$statistic)
```

## Ex 2: Mushroom gills

Estimate (make a confidence interval for) the proportion of mushrooms whose `gill.size` is broad. (For this variable, b=broad; n=narrow.) Report and interpret the interval. Use a reasonable confidence level of your choosing.

**Answer.**
A 95% confidence interval for the proportion of mushrooms whose gill size is broad is given by (0.6305843, 0.7214433). I am 95% confident that between 63.1% and 72.1% of mushrooms have broad gill size.

```{r}
prop.test(sum(mush$gill.size=='b'),n,conf.level=0.95)
```

## Ex 3: colored mushrooms
a. Look at the mushrooms for whose `cap.color` equal `w` (white). Test whether there is evidence that the proportion of these mushrooms that are edible is greater than 0.5. Use $\alpha=0.05$. 

Report the test statistic, p-value, and conclusion.

**Answer.**
There were 56 white mushrooms in the sample, out of which 41 (73.21%) were edible. To formally test whether there is evidence that the true proportion is greater than 0.05, the following hypotheses are used:

$$H_0:\pi=0.5; \quad H_A: \pi > 0.5.$$

The test statistic is $|z_0|=3.341$ and the p-value is 0.0004177. The null is rejected in favor of the alternative. The data provide strong evidence that more than 50% of white mushrooms are edible.

```{r}
white <- subset(mush, cap.color=='w')
n.white <- nrow(white)

# descriptive summary
table(white$edible)
prop.table(table(white$edible))

# hypothesis test
prop.test(sum(white$edible=='e'),n.white,p=0.5,alternative='greater')
test2 <- prop.test(sum(white$edible=='e'),n.white,p=0.5,alternative='greater')
sqrt(test2$statistic)
```

b. Compare your findings to exercise 1. Do your findings suggest any association between cap color and edibility? 

**Answer.**
In exercise 1, when looking at the full mushroom population, we found that the data provided no strong evidence that more than 50% of mushrooms are edible. In contrast, we found strong evidence that more than 50% of white mushrooms are edible. This suggests that the white mushrooms have different edibility patterns than the broader population and, thus, an association between cap color and edibility. 


## Ex 4: Writing a function (extra practice -- not required)

a. Write an R function to calculate the large-sample confidence interval for $\pi$.  Revisit Lab 1 for an example of writing a function.

**Answer.**
Answers may vary. Here are some possible approaches from DSA students.

Example 1:
```{r}
conf.int <- function(n,x,alpha)
{
  pe <- x/n
  mult <- qnorm(1-(alpha/2))
  se <- sqrt((pe*(1-pe))/n)
  moe <- mult*se
  l.bound <- pe - moe
  u.bound <- pe + moe
  ci <- c(l.bound, u.bound)
  return(ci)
}

conf.int(250,200,0.05)
```

Example 2:
```{r}
ls_ci <- function(pi.hat,alpha,n){
  z.star <- qnorm(1-(alpha/2))
  se <- sqrt((pi.hat*(1-pi.hat))/n)
  lower.interval <- pi.hat - z.star*se
  upper.interval <- pi.hat + z.star*se
  conf.interval <- c(lower.interval, upper.interval)
  print(conf.interval)
}

ls_ci(.6,.05,1000)


```

Example 3:

```{r}
ci <- function(x, n, alpha) {
  p <- x/n
  z.score <- qnorm(1 - alpha / 2, 0, 1)
  z <- sqrt((p * (1 - p)) / n) * z.score
  interval <- c(p - z, p + z)
  return(interval)
}
```
