---
title: "Lab 2"
author: "Saransh Rakshak"
date: "9/6/2024"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages and libraries

When you open an R session, a few libraries (base, stats, graphics) are already loaded.  There are many other libraries available that contain functions for statistical analysis and datasets.  CRAN is a repository from which these packages can be installed easily.  

To load a library that is already installed, use the `library()` function.  You need to do this each time you begin an R session (or at the start of an RMarkdown document that requires the library.). 

Here's an example: the dataset called `galaxies` is contained in the MASS package.  It will not load without loading the library.  Run the code below to see the error message.

```{r}
data(galaxies)
```

After you load the MASS library, it will work.

```{r}
library(MASS)
data(galaxies)  # this ''attaches'' the data set to your workspace.
head(galaxies)
hist(galaxies)  # remember the word for the shape of this distribution?
```

MASS is one of a small number of packages that are installed when R is installed.  For most other packages, first install them using the function `install.packages()`.  When R asks you to select a CRAN mirror, choose one that is geographically close to you.  After you install the package once, you can then use the package by loading the library with the `library()` function at the beginning of each session. 

Anyone can publish an R package on CRAN.  If you are using a package to perform a statistical analysis, look at the documentation and make sure the package seems to come from a reliable source. 

The code below would install and load the package "abind", which I use often to store data or simulation results in multidimensional arrays. (Optional:) Try installing and loading a package of your choosing here.  Some popular packages include "scales," which lets you control scale and color transparency in a plot, or "reshape2," which can turn wide data sets into tall ones.
```{r}
# Optional: replace "abind" in the code below with any package of your choosing and uncomment.  

#install.packages("abind")
#library(abind)
```

You can also install a package with a point-and-click GUI by choosing "Tools" in the RStudio window and clicking "Install packages".

## Displaying multiple panels in one plot
The `par` function lets you control the graphics created in R.  It has many options that control the margins, scale, and titles of plots.  The `mfrow` option allows you to arrange panels of plots in a grid.  Set `mfrow=c(a,b)` to use a grid with $a$ rows and $b$ columns.

```{r}
par(mfrow=c(1,2))
hist(rnorm(100))
boxplot(rnorm(100))
```

The following code switches the arrangement to have two rows.
```{r}
par(mfrow=c(2,1))
hist(rnorm(100))
boxplot(rnorm(100))
```

Note: The methods for displaying multiple plots are quite different if you use ggplot2.

## Displaying multiple panels in one plot: R markdown
You can also put figures side by side in an R markdown document by specifying certain options when you define the code chunk. For example, the options `fig.show="hold"` and `out.width="50%"` puts two figures side by side.

```{r, fig.show="hold", out.width="50%"}
hist(rnorm(100))
boxplot(rnorm(100))
```

## R with missing values
Some functions in R will not work in the presence of missing values.  Most plotting functions will just ignore missings, but many calculations will return the value NA. 

```{r}
# create a vector with missing values
x <- c(4,18,9,NA,0)
```

The vector `x` has an element whose value is NA, which is a special value in R representing a missing value.  If we try to use the mean function on `x`, the answer is NA.  The option `na.rm=T` can be used to ignore the missing values.

```{r}
mean(x)
mean(x, na.rm=T)  # use na.rm=T to ignore NAs
```

You might get data in which the missing values are not coded as NA values. Here, `y` has a missing value but it is not treated as NA.  Further, R thinks that `y` is a vector of character data because of the "no response" value.
```{r}
y <- c('no response',3,4,1,12)
is.na(y[1])  # returns TRUE if y[1] is NA
class(y)
```

The mean of y cannot be calculated, even with `na.rm=T`, because `y` is a character vector.

```{r}
mean(y, na.rm=T) 
```
The best way to prevent this is to ensure that R recognizes NAs when you first read in the data using the `na.strings` option in `read.csv`.  Below is one quick and dirty way to fix it.  The `as.numeric` function will turn `y` into a numeric vector.  Any elements than cannot be turned into numeric values will become NAs.  *Be careful with this approach:* it will not work as you expect on a variable whose class is `factor`, for example.

```{r}
y <- as.numeric(y)
y
class(y)
```
Missing values are addressed differently in functions of two variables, such as the `cor()` function.
The argument `use = 'complete.obs'`  will omit missing values.  See the R documentation for different options for the "use" argument.

```{r}
cor(x, y, use='complete.obs')
```

# Exercises


## Exercise 1: Explore associations in Airbnb data

Use the Airbnb data from last week, found in the file `airbnb_asheville.csv`. **Make sure to put all of your code in code chunks.**

The following R functions will help in answering the questions below: mean, median, quantile, boxplot, subset, table, prop.table. 

a. Calculate the mean, median, and standard deviation of the variable `price` among only listings for which `room_Type` is `Entire home/apt`.  Then calculate the mean, median, and standard deviation of the variable `price` among only listings for which `room_type` is `private room`.    (You did this already last week. Try using `tapply` this time!)

**Answer:**
Old Approach (R native):
```{R}
data <- read.csv("airbnb_asheville.csv")

# 'Entire home/apt'
full_house <- subset(data, room_type == "Entire home/apt")

mean(full_house$price)
median(full_house$price)
sd(full_house$price)

# 'Private room'
private_house <- subset(data, room_type == "Private room")

mean(private_house$price)
median(private_house$price)
sd(private_house$price)

```

New Approach (tapply):
```{R}

data$room_type <- as.factor(data$room_type)

tapply(data$price, data$room_type, mean)
tapply(data$price, data$room_type, median)
tapply(data$price, data$room_type, sd)

```



b. Use boxplots or histograms to display the distribution of `price` grouped by `room_Type`.  Based on the results in parts a and b, is price associated with room type?
There is a code chunk below that gives an example of syntax to make grouped boxplots.

**Answer:**
```{R}

boxplot(price ~ room_type, data = data, main = 'price by room type')
boxplot(log(price)~room_type, data=data, main ='log price by room type')

```


```{r}
# create a data frame of random numbers
randoms <- data.frame(size=rnorm(100,20,3), species=as.factor(sample(1:4,100,replace=TRUE)))

# the syntax a ~ b will make boxplots of variable "a"
# grouped by variable "b".
# in the "data=" put the name of the data frame
boxplot(size ~ species, data=randoms)
```

b. Calculate Pearson's correlation between `price` and `number_of_reviews`.  Use the `cor()` function. What does this suggest about the relationship between the variables? Make a scatterplot of `number_of_reviews` and `price`. Does their relationship look linear?

**Answer:**
```{R}

cor(data$price, data$number_of_reviews)
plot(data$price, data$number_of_reviews, xlab = 'price', ylab = 'number of reviews')

```

c. Now apply a log transformation to `price` using the `log()` function. The log transformation is often used for highly skewed variables such as price. Again make a scatterplot of  log(`number_of_reviews`+1) and log(`price`). Calculate Pearson's correlation between the two log-transformed variables. What do these two calculations suggest about the association between price and the number of reviews?


**Answer:**
```{R}

cor(log(data$price), log(data$number_of_reviews+1))
plot(log(data$price), log(data$number_of_reviews+1))

```



d. Divide the data into two subsets:  those listings with 50 or more reviews and those with less than 50 reviews.  Use boxplots or histograms to compare the distribution of `price` among the two subsets.  What do these plots suggest about the association between price and the number of reviews?

**Answer:**
```{R}

under_50 <- subset(data, number_of_reviews < 50)
over_50 <- subset(data, number_of_reviews >= 50)

par(mfrow=c(1,2))

boxplot(under_50$price, over_50$price, main = "Price by reviews")
axis(1,at=1:2,labels=c("under 50","over 50"))

boxplot(log(under_50$price), log(over_50$price), main = "Log Price by reviews")
axis(1,at=1:2,labels=c("under 50","over 50"))

```




e. Create a new categorical variable in your data set that takes one value if the listing has 50+ reviews and a different value if the listing has less than 50 reviews. Call the new variable `high_listings`. Convert the `high_listings` variable to a factor using the `as.factor` function.  (The example code below may help.)   

Then create a frequency table that shows the counts for your new variable and the `room_type`. Use row or column proportions to investigate whether `room_type` is associated with `high_listings`. Give your answer about whether the variables are associated and support it with the relevant proportions that you calculated.

**Answer: Since proportions are similar, there is no clear association that can be seen.**
```{R}

# Creating frequency table
data$high_listings <- as.factor(data$number_of_reviews >= 50)
room_type_high <- table(data$room_type, data$high_listings)
room_type_high

# Using row proportions to see association
prop.table(room_type_high)

```



```{r}
# Some example code
# First I generate some artificial data
some_data <- data.frame(x=rnorm(10),y=as.factor(sample(1:3,10,replace=T,prob=c(.1,.4,.5))))

# append to the data frame a vector called "x_positive" 
# "x_positive" indicates whether x is >0 for that observation.
some_data$x_positive <- as.factor(some_data$x > 0)

# use the table function to create a two way table
some_data_table <- table(some_data$y,some_data$x_positive)
some_data_table

# prop.table can be used to find row and column proportions
# the margin option controls whether you get row or column proportions
prop.table(some_data_table,margin=1) # row proportions
prop.table(some_data_table,margin=2)  # column proportions


```



## Exercise 3
There is a data set called `iris` in the MASS R package. Load the MASS library and the iris data. Use `?iris` to view the documentation for the data set.


a. Create four boxplots showing `Sepal.Length`, `Sepal.Width`, `Petal.Length`, `Petal.Width `grouped by `Species`. Use `par(mfrow)` to display them with a single plot with four panels.

**Answer:**
```{R}
# Loading dataset 
library(MASS)
data(iris)
head(iris)
```

```{R}
# Creating boxplots

par(mfrow=c(2,2))
boxplot(iris$Sepal.Length ~ iris$Species, xlab = 'Sepal Length', ylab = "Species")
boxplot(iris$Sepal.Width ~ iris$Species, xlab = 'Sepal Width', ylab = "Species")
boxplot(iris$Petal.Length ~ iris$Species, xlab = 'Petal Length', ylab = "Species")
boxplot(iris$Petal.Width ~ iris$Species, xlab = 'Petal Width', ylab = "Species")

```

b. Suppose you were given these measurements of an iris at random: `Sepal.Length`=6, `Sepal.Width`=2.7, `Petal.Length`=6, `Petal.Width `=1.6. Which species do you think it would belong to? Are any of its measurements surprising?

**Answer: Likely versicolor species due to Sepal Length and Sepal Width.**


c. Use some plots and summary statistics to continue exploring the data. Propose a simple rule for classifying iris species based on their petal and sepal measurements. (Your rule might look something like this: if petal.width > 5 and 2< sepal.width < 4, classify as "virginica")

**Answer: IF petal.length < 2 AND petal.width < 1: Setosa. ELIF 2 < petal.length < 4.7 AND 1 < petal.width < 1.7: Versicolor. ELSE Virginica.**

