---
title: "Week 10 Lab"
author: "name"
date: "date"
output:
  html_document: default
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

(You might want to knit the document the first time you read it to view the notation.)



## Temporarily excluding observations
Sometimes when assessing influence of outliers on an analysis, you'll want to perform an analysis that temporarily excludes one or more observation.  Here is a recap of how to do that.

First, I create a data frame of simulated data to work with.  I'm interested in using regression to predict `y` using `x`.
```{r}
# simulate a data frame with 12 observations
my_data <- data.frame(x=rnorm(12),y=rnorm(12,3,0.65))
```

Suppose I want to investigate the influence of the observation in row 9.  The syntax `[-i,]` will exclude the ith row from a data frame.  I can use this syntax in the `lm` function to exclude row 9 from the regression without creating a new data frame.

The code below finds the simple linear regression line with the full data and again with row 9 excluded.

```{r}
lm_full_data <- lm(y~x, data=my_data)
lm_no_row9 <- lm(y~x, data=my_data[-9,])
```


# Exercises

# Ex 1 (tomatoes)

Revisit the tomato data from Lab 8.  

a. Find a simple linear regression line that can predict the tomato height at measurement 2 (height2) using the pH measurement.  Report the slope and intercept of the regression line.

**Answer.**


b. Use your answer from part (a) to predict the height at measurement 2 for a tomato plant whose soil pH is 5.1.

**Answer.**

c. Find a 95\% confidence interval for the slope of the regression line.

**Answer.**

## Ex 2: (Expenditures)

 The data set `2015_revenue_expenditures.csv`  contains records of government revenue and expenditures for 150 of the largest cities in the U.S in 2015.  Among the variables measured are `Intergovt revenue`, which records the total revenue in the cities coming from state and federal government, and expenditures in various spending categories. 

a. Find the simple linear regression equation to predict expenditures in education services (`Education Services Expend`.) from  `Intergovt revenue`.  Give the estimated regression equation.  

**Answer.**

b. Perform a test of the hypotheses 

$$H_0: \beta_1 = 0; \quad H_A: \beta_1 > 0.$$

Use $\alpha=0.01$. Report the test statistic, p-value, decision, and conclusion in the context of the problem. 

**Answer.**

c. Create a plot of the x variable vs residuals.  Also create a Normal quantile plot of the residuals.  Write a sentence explaining whether the modeling assumptions of normality and equal variances appear to be reasonable for these data.

**Answer.**


d. Find the observation in the data set for VA: Chesapeake.  What is the observed value of $y_i$, the response variable, for this observation?  

**Answer.**


e.  What value does the estimated regression line from part (a) predict for this observation? 

**Answer.**


f. Calculate its residual by subtracting the predicted value from the observed value. ($\widehat{\epsilon}_i = y_i - \widehat{y}_i$.)  You can compare this to the value automatically calculated by R using the `residuals` function to check your answer.

**Answer.**



## Ex 3: (body fat)

The data set `BodyFat.csv` on Canvas contains body fat percentage and several other measurements for a sample of 253 men.  Direct measurement body fat percentage can be a cumbersome process, so it might be useful to find a different measurement that is an accurate predictor of body fat.   

a. Find a regression line that predicts `BODYFAT` from `WEIGHT`.  Write the estimated regression equation and the $R^2$ for this model.

**Answer.**


b. Now find a regression line that predicts `BODYFAT` from `ABDOMEN` (the abdomen circumference).  Write the estimated regression equation and the $R^2$ for this model.

**Answer.**


c. Based on the $R^2$ values, is weight or abdomen circumference a stronger predictor of of body fat?  (Which variable accounts for a greater proportion of the variability in body fat measurements?)

**Answer.**




# Ex 4 (more diabetes)

Revisit the diabetes data from Lab 8.  


a. Provide a descriptive plot and summary statistic that describes the relationship between glucose (glu) and diastolic blood pressure (bp).  Does there appear to be a strong association between the two variables?

**Answer.**


b. Fit a simple linear regression line to predict glucose (glu) using blood pressure (bp). Report the slope and intercept of the estimate regression line. Give a one-sentence interpretation of the value of the slope.

**Answer.**


c. Make 90% confidence intervals for the intercept and slope of the regression line. 

**Answer.**


d. Make a plot of the fitted values vs. the residuals. Do the assumptions of equal variances and linearity appear to be reasonable?

**Answer.**


e. Make a normal quantile plot of the residuals. Does the assumption of normality appear to be reasonable?

**Answer.**

